Look-Ahead Screening Rules for the Lasso

### 2021-06-24 paper/main.tex
Revise paper based on reviewer comments

\author[1]{Johan Larsson%
  \thanks{%
    \href{mailto:johan.larsson@stat.lu.se}{\url{johan.larsson@stat.lu.se}}
  }%
}
\affil[1]{The Department of Statistics, Lund University}

@@

\begin{abstract}
  The lasso is a popular method to induce shrinkage and sparsity in the
  solution vector (coefficients) of regression problems, particularly
  when there are many predictors relative to the number of observations. Solving
  the lasso in this high-dimensional setting can, however, be computationally
  demanding. Fortunately, this demand can be alleviated via the
  use of \emph{screening rules} that discard predictors prior to
  fitting the model, leading to a reduced problem to be solved.
  In this paper, we present a new screening strategy: \emph{look-ahead
    screening}. Our method uses safe screening rules to find a range of
  penalty values for which a given predictor cannot enter the model,
  thereby screening predictors along the remainder of the path.
  In experiments we show that these look-ahead screening rules improve the
  performance of existing screening strategies.
\end{abstract}

@@

Let \(X \in \mathbb{R}^{n \times p}\) be the design matrix with \(n\)
observations and \(p\) predictors
and \(y \in \mathbb{R}^n\) the response vector.
The lasso is represented by the following convex optimization problem:
\begin{equation}
  \label{eq:primal}
  \operatorname*{minimize}_{\beta \in \mathbb{R}^p} \left\{P(\beta; \lambda)
  = \frac 1 2 \lVert y - X\beta \rVert_2^2 + \lambda \lVert \beta \rVert_1\right\}
\end{equation}
where \(P(\beta;\lambda)\)
is the \emph{primal} objective. We let \(\hat \beta_\lambda\) be the solution to
\eqref{eq:primal} for a given \(\lambda\) and \(\tilde
\beta_\lambda\) an estimate of \(\hat\beta_\lambda\).\footnote{Such as
  the current estimate in an iterative solver tasked with solving
  \eqref{eq:primal}.}

% The Karush--Kuhn--Tucker (KKT) stationarity condition for \eqref{eq:primal}
% holds that
% \begin{equation}
%   \label{eq:stationarity}
%   0 \in X^T(X\beta - y) + \lambda\partial,
% \end{equation}
% where \(\partial\) is the subdifferential of the \(\ell_1\)-norm, with
% its \(j\)th element corresponding to
% \begin{equation}
%   \label{eq:subdifferential}
%   \partial_j \in
%   \begin{cases}
%     \{\sign(\beta_j)\} & \text{if } \beta_j \neq 0 \\
%     [-1,1]             & \text{otherwise.}
%   \end{cases}
% \end{equation}
Moreover, the dual problem of \eqref{eq:primal} is
\begin{equation}
  \label{eq:dual}
  \operatorname*{maximize}_{\theta \in \mathbb{R}^n} \left\{ D(\theta; \lambda) =
  \frac 12 y^T y - \frac{\lambda^2}{2} \left\lVert \theta
  - \frac y \lambda \right\rVert_2^2\right\}
\end{equation}
where \(D(\theta; \lambda)\) is the \emph{dual} objective.
The relationship between the primal and dual problems is given by
\(y = X\hat\beta_\lambda + \lambda \hat\theta_\lambda.\)

@@

Suppose, now, that we have solved the lasso for \(\lambda\); then
for any given \(\lambda^* \leq \lambda\);
the Gap Safe rule~\citep{ndiaye2017} discards the \(j\)th
predictor if
\begin{equation}
  \label{eq:gap-safe-rule}
  |X^T \tilde \theta_\lambda|_j + \lVert x_j\rVert_2
  \sqrt{\frac{1}{\lambda_*^2}
    G(\tilde \beta_\lambda, \tilde \theta_\lambda; \lambda^*)}
  < 1
\end{equation}
where
\[
  \tilde\theta_\lambda = \frac{y - X\tilde\beta_\lambda}{
    \max\big( |X^T(y - X\tilde\beta_\lambda)|, \lambda\big)}
\]
is a dual-feasible point~\cite{ndiaye2017} obtained through dual scaling.

Next, observe that \eqref{eq:gap-safe-rule} is a quadratic inequality with
respect to \(\lambda_*\), which
means that it is trivial to discover the boundary points via the
quadratic formula:
\[
  \lambda_* = \frac{-b \pm \sqrt{b^2 - 4ac}}{2a}
\]
where
\[
  \begin{aligned}
    a & = \big( 1 - | x_j^T \tilde\theta_\lambda|\big)^2 -
    \frac 12 \tilde\theta_\lambda^T \tilde\theta_\lambda \lVert x_j\rVert_2^2      \\
    b & = \big(\tilde\theta_\lambda^T y - \lVert \tilde\beta_\lambda \rVert_1\big)
    \lVert x_j \rVert_2^2                                                          \\
    c & = - \frac 12 \lVert y - X\tilde\beta_\lambda\rVert_2^2
    \lVert x_j\rVert_2^2.                                                          \\
  \end{aligned}
\]
By restricting ourselves to an index \(j\) corresponding to a predictor that is
inactive at \(\lambda\) and recalling that we have \(\lambda_* \leq \lambda\) by
construction, we can inspect the signs of \(a\), \(b\), and \(c\) and find a
range \(\lambda\) values for which predictor \(j\) must be inactive. Using this
idea for the lasso path---a grid of \(\lambda\) values starting from the null
(intercept-only) model, which corresponds to \(\lambda_\text{max} =
\max_i|x_i^Ty|\), and finishing at fraction of this (see
\autoref{seq:simulations} for specifics)---we can screen predictor \(j\) for all
upcoming \(\lambda\)s, possibly discarding it for multiple steps on the path
rather than just the next step. We call this idea \emph{look-ahead screening}.

To illustrate the effectiveness of this screening method, we consider an
instance of employing look-ahead screening for fitting a full lasso path to the
\emph{leukemia} dataset~\citep{golub1999}.  At the first step of the path, the
screening method discards 99.6\% of the predictors for the steps up to and
including step 5. The respective figures for steps 10 and 15 are 99.3\% and
57\%.  At step 20, however, the rule does not discard a single predictor.  In
\autoref{fig:case-study}, we have visualized the screening performance of
look-ahead screening for a random sample of 25 predictors from this dataset.

\begin{figure}[hbtp]
  \centering
  \includegraphics{figures/casestudy.pdf}
  \caption{This figure shows the predictors screened at the first step of the
    lasso path via look-ahead screening for a random sample of
    20 predictors from the \emph{leukemia}
    dataset. A blue square indicates that the corresponding predictor can
    be discarded at the respective step.\label{fig:case-study}}
\end{figure}

@@

To fit the lasso, we
use cyclical coordinate descent~\cite{friedman2010}.
We consider the solver to have converged whenever
the duality gap as a fraction of the primal value
for the null model is less than or equal to \(10^{-6}\) and the amount of
\emph{infeasibility}, which we define as \(\max_i\big( |c_i| -
\lambda\big)\), as a fraction of \(\lambda_\text{max}\) is lower than or
equal to \(10^{-5}\).

The code used in these experiments is programmed in \textsf{C++}.
We use a combination of the \textsf{renv} R
package and Singularity to set up a
reproducible environment for the experiments. All
source code
can be found at \url{https://github.com/jolars/LookAheadScreening/}. An HPC
cluster node with two Intel Xeon E5-2650 v3 processors (Haswell, 20 compute
cores per node) and 64 GB of RAM was used to run the experiments.

We
run experiments on a design with \(n = 100\) and \(p = 50\,000\),
drawing the rows of \(X\) i.i.d.
from \(\mathcal{N}(0, \Sigma)\) and \(y\)
from \(\mathcal{N}(X\beta, \sigma^2I)\) with \(\sigma^2 =
\beta^T\Sigma\beta/\text{SNR}\), where SNR is the signal-to-noise ratio.  We set
\(5\) coefficients, equally spaced throughout the coefficient vector, to 1 and
the rest to zero. Taking inspriation from \citet{hastie2020}, we
consider SNR values of 0.1, 1, and 6.

@@

\begin{figure}[hbtp]
  \centering
  \includegraphics{figures/simulateddata-timings.pdf}
  \caption{Standard box plots of timings to fit a full lasso path to
    a simulated dataset with \(n = 100\), \(p = 50\,000\), and five true
    signals.}
  \label{fig:simulated-data}
\end{figure}

\section{Discussion}\label{sec:discussion}

In this paper, we have presented \emph{look-ahead screening}, which is a
novel method to screen predictors for a range of penalization values along
the lasso regularization path using Gap Safe screening. Our results show that
this type of screening can yield considerable improvements in performance.

The idea is general and can therefore be extended to any type of safe screening
rule and also used in tandem with heuristic screening rules in order to avoid
expensive KKT computations.

>>>>

\author[1]{Johan Larsson%
  \thanks{%
    Corresponding author: \href{mailto:johan.larsson@stat.lu.se}{\url{johan.larsson@stat.lu.se}}
  }%
}
\affil[1]{The Department of Statistics, Lund University}

@@

\begin{abstract}
  The lasso is a popular method to induce shrinkage and sparsity in the
  solution vector (coefficients) of regression problems, particularly
  when there are many predictors relative to the number of observations. Solving
  the lasso in this high-dimensional setting can, however, be computationally
  demanding. Fortunately, this demand can be alleviated via the
  use of \emph{screening rules} that discard predictors prior to
  fitting the model, leading to a reduced problem to be solved.
  In this paper, we present a new screening strategy: \emph{look-ahead
    screening}. Our method uses safe screening rules to find a range of
  penalty values for which a given predictor cannot enter the model,
  thereby screening predictors along the remainder of the path.
  In experiments we show that these look-ahead screening rules outperform the
  active warm-start version of the Gap Safe rules.
\end{abstract}

@@

Let \(X \in \mathbb{R}^{n \times p}\) be the design matrix with \(n\)
observations and \(p\) predictors
and \(y \in \mathbb{R}^n\) the response vector.
The lasso is represented by the following convex optimization problem:
\begin{equation}
  \label{eq:primal}
  \operatorname*{minimize}_{\beta \in \mathbb{R}^p} \left\{P(\beta; \lambda)
  = \frac 1 2 \lVert y - X\beta \rVert_2^2 + \lambda \lVert \beta \rVert_1\right\}
\end{equation}
where \(P(\beta;\lambda)\)
is the \emph{primal} objective. We let \(\hat \beta_\lambda\) be the solution to
\eqref{eq:primal} for a given \(\lambda\).
Moreover, the dual problem of \eqref{eq:primal} is
\begin{equation}
  \label{eq:dual}
  \operatorname*{maximize}_{\theta \in \mathbb{R}^n} \left\{ D(\theta; \lambda) =
  \frac 12 y^T y - \frac{\lambda^2}{2} \left\lVert \theta
  - \frac y \lambda \right\rVert_2^2\right\}
\end{equation}
where \(D(\theta; \lambda)\) is the \emph{dual} objective.
The relationship between the primal and dual problems is given by
\(y = X\hat\beta_\lambda + \lambda \hat\theta_\lambda.\)

@@

Suppose, now, that we have solved the lasso for \(\lambda\); then
for any given \(\lambda^* \leq \lambda\);
the Gap Safe rule~\citep{ndiaye2017} discards the \(j\)th
predictor if
\begin{equation}
  \label{eq:gap-safe-rule}
  |X^T \theta_\lambda|_j + \lVert x_j\rVert_2
  \sqrt{\frac{1}{\lambda_*^2}
    G(\beta_\lambda, \theta_\lambda; \lambda^*)}
  < 1
\end{equation}
where
\[
  \theta_\lambda = \frac{y - X\beta_\lambda}{
    \max\big( |X^T(y - X\beta_\lambda)|, \lambda\big)}
\]
is a dual-feasible point~\cite{ndiaye2017} obtained through dual scaling.

Observe that \eqref{eq:gap-safe-rule} is a quadratic inequality with
respect to \(\lambda_*\), which
means that it is trivial to discover the boundary points via the
quadratic formula:
\[
  \lambda_* = \frac{-b \pm \sqrt{b^2 - 4ac}}{2a} \quad \text{where} \quad
  \begin{aligned}
    a & = \big( 1 - | x_j^T \theta_\lambda|\big)^2 -
    \frac 12 \theta_\lambda^T \theta_\lambda \lVert x_j\rVert_2^2,      \\
    b & = \big(\theta_\lambda^T y - \lVert \beta_\lambda \rVert_1\big)
    \lVert x_j \rVert_2^2,                                                          \\
    c & = - \frac 12 \lVert y - X\beta_\lambda\rVert_2^2
    \lVert x_j\rVert_2^2.                                                          \\
  \end{aligned}
\]
By restricting ourselves to an index \(j\) corresponding to a predictor that is
inactive at \(\lambda\) and recalling that we have \(\lambda_* \leq \lambda\) by
construction, we can inspect the signs of \(a\), \(b\), and \(c\) and find a
range \(\lambda\) values for which predictor \(j\) must be inactive. Using this
idea for the lasso path---a grid of \(\lambda\) values starting from the null
(intercept-only) model, which corresponds to \(\lambda_\text{max} =
\max_i|x_i^Ty|\), and finishing at fraction of this (see
\autoref{seq:simulations} for specifics)---we can screen predictor \(j\) for all
upcoming \(\lambda\)s, possibly discarding it for multiple steps on the path
rather than just the next step. We call this idea \emph{look-ahead screening}.

To illustrate the effectiveness of this screening method, we consider an
instance of employing look-ahead screening for fitting a full lasso path to the
\emph{leukemia} data set~\citep{golub1999}.  At the first step of the path, the
screening method discards 99.6\% of the predictors for the steps up to and
including step 5. The respective figures for steps 10 and 15 are 99.3\% and
57\%.  At step 20, however, the rule does not discard a single predictor.  In
\autoref{fig:case-study}, we have visualized the screening performance of
look-ahead screening for a random sample of 25 predictors from this data set.

\begin{figure}[hbtp]
  \centering
  \includegraphics{figures/casestudy.pdf}
  \caption{This figure shows the predictors screened at the first step of the
    lasso path via look-ahead screening for a random sample of
    20 predictors from the \emph{leukemia}
    data set. A blue square indicates that the corresponding predictor can
    be discarded at the respective step.\label{fig:case-study}}
\end{figure}

@@

To fit the lasso, we
use cyclical coordinate descent~\cite{friedman2010}.
We consider the solver to have converged whenever
the duality gap as a fraction of the primal value
for the null model is less than or equal to \(10^{-6}\) and the amount of
\emph{infeasibility}, which we define as
\(\max_j\big( |x_j^T(y - X\beta_\lambda)| -\lambda\big)\), as a
fraction of \(\lambda_\text{max}\) is lower than or
equal to \(10^{-5}\).

Source code for the experiments, including a container to facilitate
reproducibility, can be found at
\url{https://github.com/jolars/LookAheadScreening/}. An HPC
cluster node with two Intel Xeon E5-2650 v3 processors (Haswell, 20 compute
cores per node) and 64 GB of RAM was used to run the experiments.

We run experiments on a design with \(n = 100\) and \(p = 50\,000\),
drawing the rows of \(X\) i.i.d. from \(\mathcal{N}(0, \Sigma)\) and \(y\)
from \(\mathcal{N}(X\beta, \sigma^2I)\) with \(\sigma^2 =
\beta^T\Sigma\beta/\text{SNR}\), where SNR is the signal-to-noise ratio.  We set
\(5\) coefficients, equally spaced throughout the coefficient vector, to 1 and
the rest to zero. Taking inspiration from \citet{hastie2020}, we
consider SNR values of 0.1, 1, and 6.

@@

\begin{figure}[hbtp]
  \centering
  \includegraphics{figures/simulateddata-timings.pdf}
  \caption{Standard box plots of timings to fit a full lasso path to
    a simulated data set with \(n = 100\), \(p = 50\,000\), and five true
    signals.}
  \label{fig:simulated-data}
\end{figure}

\section{Discussion}%
\label{sec:discussion}

In this paper, we have presented \emph{look-ahead screening}, which is a novel
method to screen predictors for a range of penalization values along the lasso
regularization path using Gap Safe screening. Our results show that this type of
screening can yield considerable improvements in performance for the standard
lasso. For other loss functions,
\eqref{eq:gap-safe-rule} may no longer reduce to a quadratic inequality and will
hence require more computation. Nevertheless, we believe
that applying these rules in these cases is feasible and likely to result
in comparable results.

Moreover, the idea is general and can therefore be extended to any type of safe
screening rule and also used in tandem with heuristic screening rules in order
to avoid expensive KKT computations. Finally, although we only cover one type of
solver in our experiments, note that our screening method is
agnostic to the solver used to solve the lasso and that we therefore expect our
results to be generalizable.
